`sdmTMB`: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-10-18 doc/paper.Rmd
Add example

```{r, echo = FALSE, include = FALSE, message = FALSE, warnings = FALSE}
library(here)
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warnings = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

@@

We use the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time.
The general form of a GLMM is 
$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$. 
<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
The full `sdmTMB` model is
$$
g(\mu_{s,t}) = 
\bm{X}_{1,\bm{s},t} \bm{\beta} + 
\bm{\alpha}_g +
\bm{X}_{2,\bm{s},t} \gamma_t + 
\bm{X}_{3,\bm{s},t} \zeta_{\bm{s}} +
\omega_{\bm{s}} + 
\epsilon_{\bm{s},t},
$$
where

@@

>>>>

```{r, echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
library(here)
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

@@

We use the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time.
The general form of a GLMM is 
$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$. 
<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
The full `sdmTMB` model is
$$
g(\mu_{\bm{s},t}) = 
\bm{X}_{1,\bm{s},t} \bm{\beta} + 
\bm{\alpha}_g +
\bm{X}_{2,\bm{s},t} \gamma_t + 
\bm{X}_{3,\bm{s},t} \zeta_{\bm{s}} +
\omega_{\bm{s}} + 
\epsilon_{\bm{s},t},
$$
where

@@

```{r libs, echo = TRUE}
library(sdmTMB)
library(ggplot2)
```

```{r}
theme_set(ggsidekick::theme_sleek())
```

First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations. The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between v vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in (examples included in `?make_mesh`).
For example, an mesh with a cutoff distance of 10km can be made with

```{r mesh, echo = TRUE, cache=FALSE}
mesh <- make_mesh(pcod_2011, xy_cols = c("X", "Y"), cutoff = 10)
```

```{r plot-mesh, echo = TRUE, fig.height=5, fig.width=5, out.width="3.25in"}
plot(mesh)
```

```{r fit-pcod, cache=TRUE, echo = TRUE}
fit <- sdmTMB(density ~ 0 + as.factor(year) + s(depth), data = pcod_2011,
 time = "year", spde = mesh, family = tweedie(link = "log")
)
```

```{r print-fit, echo = TRUE}
print(fit)
```

```{r tidy, eval=TRUE, echo=TRUE}
tidy(fit, conf.int = TRUE)
```

```{r, eval=TRUE, echo=TRUE}
tidy(fit, effects = "ran_pars", conf.int = TRUE)
```

```{r predict, cache=TRUE, echo = TRUE}
pred <- predict(fit)
```

```{r qqnorm, cache=TRUE, fig.width=4, fig.height=4, out.width="3.5in", echo = TRUE}
pred$resid <- residuals(fit)
# qqnorm(pred$resid);qqline(pred$resid) # not shown for space
```

```{r plot-resids, echo = FALSE, eval=FALSE}
ggplot(subset(pred, year == 2017), aes(X, Y, colour = resid)) +
    geom_point() + facet_wrap(~year) +
    coord_fixed() + scale_colour_gradient2()
```

```{r predict-newdata, cache=TRUE, echo = TRUE}
qcs_grid_2011 <- subset(qcs_grid, year >= 2011)
pred_qcs <- predict(fit, newdata = qcs_grid_2011)
```

```{r plot-predictions, echo = TRUE, fig.height=5, fig.width=5, out.width="4.5in", cache=TRUE}
ggplot(pred_qcs, aes(X, Y, fill = exp(est))) +
    geom_raster() + facet_wrap(~year) +
    coord_fixed() + scale_fill_viridis_c(trans = "sqrt")
```

```{r marginal, cache=TRUE, echo = TRUE, fig.height=2.5, fig.width=4, out.width="3.5in", echo = TRUE}
nd <- data.frame(depth = seq(50, 300, length.out = 100), year = 2017L)
pred_depth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = ~ 0)
ggplot(pred_depth, aes(depth, exp(est),
  ymin = exp(est - 1.96 * est_se), ymax = exp(est + 1.96 * est_se))) +
  geom_line() + geom_ribbon(alpha = 0.4) + 
  ylab("Expected density\n(kg/TODO)") +
  xlab("Depth (m)")
```

```{r sims, cache=TRUE, echo=TRUE}
pred_sims <- predict(fit, newdata = qcs_grid_2011, sims = 500)
dim(pred_sims)
```

```{r sims-cv-plot, cache=FALSE, fig.height=4, fig.width=4, out.width="3.5in", echo=TRUE}
qcs_grid_2011$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
ggplot(subset(qcs_grid_2011, year == 2011), aes(X, Y, fill = cv)) +
    geom_raster() + facet_wrap(~year) +
    coord_fixed() + scale_fill_viridis_c(trans = "log10")
```

```{r index, cache=TRUE, fig.height=2.5, fig.width=4, out.width="3.5in"}
qcs_grid_2011$area <- 4 # all 2 x 2km
pred2 <- predict(fit, newdata = qcs_grid_2011, return_tmb_object = TRUE, 
  area = qcs_grid_2011$area)
ind <- get_index(pred2)
ggplot(ind, aes(year, est)) + geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  ylab("Biomass (kg)")
```
